---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---

## Loading and preprocessing the data  
```{r loaddata,echo=TRUE, results='hide',warning=FALSE, message=FALSE}
zip <- "activity.zip"
filename <- "activity.csv"
unzip(zip,files = c(filename))
df <- read.csv(filename)
```
Data file *`r filename`* is extracted from archive *`r zip`* and the data are loaded as shown in code above.  
Per assignment instruction, data is comprised of complete records only for initial processing. (Days/intervals with no number of steps recorded are not considered in the summary data.)  
```{r tidydata,echo=TRUE, results='hide',warning=FALSE, message=FALSE}
# create working data set omitting rows with NA
dfc <- na.omit(df)
# prepend necessary 0's to interval and convert to factor
intvl_char <- function(intvl) {
  intvl <- as.character(intvl)
  require(stringr)
  intvl <- paste(str_dup("0", 4 - str_length(intvl)), intvl,sep = "")
  return(intvl)
}
dfc$interval <- factor(intvl_char(dfc$interval))
```
####Setting interval data as formatted character factor####
It has been noted that "interval" values in source data are encoded as enjambed hour and minute (i.e., 1745 indicates 5:45 PM), and are loaded by default as integer values. Treating this value as integer series for plotting introduces artificial gaps in the data (12 intervals with data at minutes 00 to 55 followed by 8  missing intervals for interval 60 through 95 in supposed integral series). Interval data values are therefore converted to factors and standardized as 4-character strings with leading 0s (for display on plot) with sequential numeric values supporting time series plot (evenly spaced values on x-axis). 

##What is mean total number of steps taken per day?##
Summarize steps data to get daily totals, then calculate mean and median.
```{r calculatemeanandmedian, message=FALSE}
require(plyr)
daily_steps<-ddply(dfc,.(date),summarize,sum(steps))
colnames(daily_steps)<-list("Date","Steps")
step_mean<-format(mean(daily_steps[["Steps"]]), scientific=FALSE)
step_median<-format(median(daily_steps[["Steps"]]), scientific=FALSE)
```
**The mean total steps taken per day is `r step_mean`.**  
**The median total steps taken per day is `r step_median`.**  
  
The histogram below plots frequency of daily steps totals reported in 2500-step ranges. NOTE: Total days (sum of frequencies) is less than 61 days because days that had no record of steps have been excluded.
```{r dailystepshistogram, fig.path="figure/",message=FALSE}
require(ggplot2)
g<-ggplot(data=daily_steps, aes(Steps)) 
g<-g + geom_histogram(binwidth=2500, color="white", fill="blue")
g + labs(y="Number of Days", x="Total Steps", title=" Frequency of Daily Steps Totals")

```

## What is the average daily activity pattern? 
Steps are grouped by interval and averaged (mean) so that mean number of steps can be plotted for each 5-minute interval across the day. Higher mean steps count indicates increased 'activity' level for that time period. 
```{r plotactivitypattern,  fig.path="figure/"}
interval_mean<-ddply(dfc,.(interval),summarize,mean=mean(steps))
colnames(interval_mean)<-list("Interval","Steps")
maxstepinterval <- as.character( interval_mean[interval_mean$Steps == max(interval_mean$Steps), ]$Interval)

g<-ggplot(data=interval_mean,aes(as.numeric(Interval),Steps))
g<-g+scale_x_discrete(limits = c(0:288) ,breaks=seq(0,288,24), labels=seq(0000,2400,200)) 
g<-g+geom_line() 
g+labs(y="Steps",x="Interval Time",title="Average Daily Activity Pattern")

```
  
**The 5-minute time interval with the highest average step count across all days begins at *`r maxstepinterval`*.** NOTE: Interval is displayed as 'military time' with the first two digits showing hours (00-23) and the second two digits showing minutes (00-55) identifying interval.  

## Imputing missing values
The first part of the assignment required excluding records with missing steps data. Checking the count of excluded rows shows a significant number of records skipped.
```{r countna}
missingsteps <- nrow(df[!complete.cases(df),])
```
**There are *`r missingsteps`* rows missing steps data that were not included in preceding analysis.**

A new working dataset can be created from original data load to include records across all time intervals for all 61 days of the experiment. A missing steps count can be filled in with the value calculated as the mean of existing values for the corresponding interval and day of the week. No rows or records will be omitted due to missing data.
```{r mutateNA}
# Add a column for day of the week so that we can calculate mean by time interval and day of week
dfc<-mutate(df,weekday=weekdays(strptime(df$date,"%Y-%m-%d")))
# create a data frame and lookup function to support updating NA with mean value for specified day of week and interval 
m_idow<-ddply(dfc,.(interval,weekday),summarize,steps=mean(steps,na.rm=TRUE))
lookup<-function(intvl,wday,stp){
  if (is.na(stp)){
    val<-round(m_idow[m_idow$interval==intvl & m_idow$weekday==wday,]$steps)
  } else { 
    val<-stp
  }
  return(val)
}
dfc<-ddply(dfc,.(date,interval,weekday),mutate,steps=lookup(interval,weekday,steps))
```  
Analysis above, which skipped days/intervals with missing data, is repeated here with augmented dataset. Mean and median values are calculated against the larger dataset.
```{r repeatanalysis}
daily_steps<-ddply(dfc,.(date),summarize,sum(steps))
colnames(daily_steps)<-list("Date","Steps")

step_mean2<-format(mean(daily_steps[["Steps"]]), scientific=FALSE)
step_median2<-format(median(daily_steps[["Steps"]]), scientific=FALSE)

```  
**The mean total steps taken per day is now calculated as *`r step_mean2`*.** (Compared to *`r step_mean`* before NA values were filled in with imputed values.)

**The median total steps taken per day is now calculated as *`r step_median2`*.** (Compared to *`r step_median`* before NA values were filled in with imputed values.)

**Adding imputed values has increased both the mean and median values for daily step totals.**

The histogram below plots frequency of new daily steps totals. Totals on some bars has increased reflecting the addition of days which had been previously excluded because they had no data. Total of all frequencies should now be 61.  

```{r repeatdailystepshistogram, fig.path="figure/"}
g<-ggplot(data=daily_steps, aes(Steps)) 
g<-g + geom_histogram(binwidth=2500, color="white", fill="blue") 
g + labs(y="Number of Days", x="Total Steps", title="Daily Total Step Frequency (with imputed data)")
```

## Are there differences in activity patterns between weekdays and weekends?
Using the augmented data and day of week, we can plot activity patterns for weekend versus weekday. Again, the interval as factor provides the time series value on the x-axis.
```{r analyzeforweekends, fig.path="figure/"}
dfc$interval <-
  factor(paste(str_dup("0",4 - str_length(
  as.character(dfc$interval)
  )),as.character(dfc$interval),sep = ""))
  dfc <-
  ddply(dfc,.(steps,date,interval,weekday),mutate,weekpart = factor(ifelse(
  weekday %in% c("Saturday","Sunday"),"weekend","weekday"
  )))
  interval_mean <-
  ddply(dfc,.(interval,weekpart),summarize,mean = mean(steps))
  colnames(interval_mean) <- list("Interval","WeekPart","Steps")
  
  g<-ggplot(data = interval_mean,aes(as.numeric(Interval),Steps)) 
  g<-g + scale_x_discrete(
  limits = c(0:288) ,
  breaks = seq(0,288,24),
  labels = seq(0000,2400,200)) 
  g<-g + geom_line() + facet_grid(WeekPart ~ .)
  g + labs(y = "Steps",
         x = "Interval Time",
         title = "Weekend vs Weekday Activity Patterns")
```
